---
title: "Sentiment analysis twitter"
author: "Miriam Lim Leal"
date: "2026-03-26"
output: html_document
---

## Sentiment analysis Twitter

```{r}
library(tidyverse)
library(tidytext)
library(textdata)
library(emoji)
library(readxl)
library(stringi)
library(syuzhet)
library(udpipe)
```

### 1. Load data

```{r}
twitter_data <- read.csv("Data/tweets_political_parties_es.csv", colClasses = c(id = "character"))
```

### 2. Clean tweet text

#### General pipeline

Remove URLs, mentions and normalise whitespace

```{r}
clean_general <- function(text) {
  
  # Remove URLs
  text <- str_remove_all(text, "https?://\\S+")
  
  # Remove mentions
  text <- str_remove_all(text, "@\\w+")
  
  # Normalize whitespace
  text <- str_squish(text)
  
  return(text)
}
```

```{r}
twitter_transformer <- twitter_data |> 
  mutate(text_clean = map_chr(tweet, clean_general))
```

### 3. Lexicon-based model

#### 3.1 Emoji mapping in Spanish

##### Load the Spanish emoji mapping (Emoji-SP).

```{r}
emoji_map <- read_excel("emoji_mapping_es.xlsx") |> 
  mutate(definition_lower = str_to_lower(definition)) |> 
  select(emoji, definition_lower, val)
```

##### Function replace emojis with text using Emoji-SP

```{r}
# This replaces the emoji symbol with the Spanish definition from Emoji-SP file
replace_emoji_with_def <- function(text, lexicon) {
  # We use stri_replace_all_fixed to swap emoji symbols for definitions
  # We add spaces and a prefix 'EMOJI_' to help the model distinguish it
  new_text <- stri_replace_all_fixed(
    text, 
    lexicon$emoji, 
    paste0(" ", lexicon$definition_lower, " "), 
    vectorize_all = FALSE
  )
  return(new_text)
}
```

##### Create a new column with emoji replaced text called emoji_text

```{r}
twitter_lexicon <- twitter_transformer %>%
  mutate(emoji_text = sapply(text_clean, replace_emoji_with_def, lexicon = emoji_map))
```

#### 3.2 Lexicon-specific cleaning pipeline

clean data (specifically column emoji_text): drop hashtags symbols, remove punctuation and make everything lowercase.

```{r}
clean_lexicon <- function(text) {
  
  # Keep hashtag words, drop the symbol
  text <- str_replace_all(text, "#(\\w+)", "\\1")
  
  # Remove all punctuation
  text <- str_remove_all(text, "[!\"#$%&'()*+,\\-./:;<=>?@\\[\\\\\\]^_`{|}~¡¿]")
  
  # Lowercase for lexicon matching
  text <- str_to_lower(text)
  
  # Normalize whitespace again after removals
  text <- str_squish(text)
  
  return(text)
}
```

```{r}
twitter_lexicon <- twitter_data |> 
  mutate(
    # Step 1: shared cleaning (URLs, mentions, whitespace)
    text_clean = map_chr(tweet, clean_general),
    
    # Step 2: replace emojis with Spanish definitions (applied to text_clean)
    emoji_text        = map_chr(text_clean, replace_emoji_with_def, lexicon = emoji_map),
    
    # Step 3: lexicon cleaning (hashtags, punctuation, lowercase)
    clean_lexicon_text = map_chr(emoji_text, clean_lexicon)
  ) |>
  select(-emoji_text)    
```

Save this as a csv

```{r}
write.csv(twitter_lexicon, "Data/twitter_lexicon_and_transformer_clean.csv", row.names = FALSE)
```

#### 3.3. Lexicon model: NRC

##### Run NRC

```{r}
nrc_results <- get_nrc_sentiment(twitter_lexicon$clean_lexicon_text, language = "spanish")
```

```{r}
# Bind the id and tweet columns from twitter_lexicon to the NRC results
nrc_results <- bind_cols(
  twitter_lexicon |> select(id, political_party, tweet),
  nrc_results
)
```

##### Save the results of NRC

```{r}
write.csv(nrc_results, "Data/NRC_results.csv", row.names = FALSE)
```

### 4. Trasformer-based model: RoBERTuito

#### 4.1 Open clean data

```{r}
clean_data <- read.csv("Data/twitter_lexicon_and_transformer_clean.csv", colClasses = c(id = "character"))
```

#### 4.2 Install prerequisites

```{r}
#install.packages("reticulate")
library(reticulate)

# Install Python dependencies:
# py_install(c("transformers", "torch", "pysentimiento"), pip = TRUE)
```

```{r}
pysentimiento <- import("pysentimiento")
```

#### 4.3 Create analyzer for sentiment

```{r}
analyzer <- pysentimiento$create_analyzer(task = "sentiment", lang = "es")
```

#### 4.4 Apply to data

```{r}
results <- lapply(clean_data$text_clean, function(x) {
  pred <- analyzer$predict(x)
  data.frame(
    label = pred$output,
    POS   = pred$probas[["POS"]],
    NEG   = pred$probas[["NEG"]],
    NEU   = pred$probas[["NEU"]]
  )
})
```

#### 4.5 Create 1 single dataframe:

```{r}
results_df <- do.call(rbind, results)
```

#### 4.6 join results to original dataframe

Only include id, political_party and tweet from "clean_data" and bind with the results from RoBERTuito

```{r}
robertuito_results <- cbind(clean_data[, c("id", "political_party", "tweet")], results_df)
```

#### 4.7 Save results

```{r}
write.csv(robertuito_results, "Data/RoBERTuito_results.csv", row.names = FALSE)
```


### 5. LLM model: Gemini 

Gemini 2.5 Flash Lite model for the analysis

- Batching: Group data into batches of 10 to 20 rows and ask Gemini to return a structured JSON list of results.
- Structured Output: Force Gemini to return only valid JSON
- Rate Limit Buffering: Implement a small Sys.sleep() time delay between batches.

#### 5.1 Setup

```{r libraries}
library(httr2)
library(tidyverse)
library(jsonlite)

```

#### 5.2 Prepare data 

Read data and select only the columns needed

```{r prepare-data}
my_data <- read.csv("Data/twitter_lexicon_and_transformer_clean.csv", colClasses = c(id = "character"))

my_data <- my_data |> select(id, political_party, text_clean)

# Normalise quotes to prevent JSON breaking
my_data <- my_data %>%
  mutate(text_clean = str_replace_all(text_clean, '[\u201C\u201D\u201E\u201F]', '"'),
         text_clean = str_replace_all(text_clean, '"', "'"))

# Define batch size (e.g., 20 rows per API call)
batch_size <- 20
my_data <- my_data %>% 
  mutate(batch_id = ceiling(row_number() / batch_size))
```

#### 5.3 Analysis Loop

```{r gemini-loop}

API_KEY <- Sys.getenv("GEMINI_KEY") #API key is saved in .Renviron file as: GEMINI_API_KEY=your_key, call usethis::edit_r_environ() to open that file and save it.

if (API_KEY == "") stop("GEMINI_KEY not found. Check your .Renviron file.")

# Select model: gemini-2.5-flash-lite, which is designed for simple classification tasks (gemini-2.5-flash is much more expensive) 
url <- paste0(
  "https://generativelanguage.googleapis.com/v1beta/models/",
  "gemini-2.5-flash-lite:generateContent?key=", API_KEY
)

# Initialise results storage and failure log
results_list  <- list()
failed_batches <- c()   # track failed batch IDs in memory
total_batches  <- max(my_data$batch_id)

for (i in 1:total_batches) {

  # Skip this batch if it was already processed.
  # This prevents silent duplicates if the loop is re-run after a crash.
  if (as.character(i) %in% names(results_list)) {
    message(sprintf("Batch %d already processed — skipping.", i))
    next
  }

  current_batch <- my_data %>% filter(batch_id == i)

  # jsonlite::toJSON handles all quote escaping internally — do not pre-escape.
  batch_json_string <- jsonlite::toJSON(
    current_batch %>% select(id, text_clean),
    auto_unbox = TRUE
  )

  prompt <- paste0(
    "You are a strict data processing tool. Do sentiment analysis on the ",
    "text data provided in JSON format below following these instructions:\n\n",
    "1. Sentiment Classification: In English, classify each tweet exactly as ",
    "'positive', 'negative', or 'neutral' (choose only one label per tweet).\n",
    "2. Explanation: Provide a concise and specific explanation for how you ",
    "decided on that label.\n\n",
    "CRITICAL: Return ONLY a valid JSON array of objects containing exactly ",
    "three keys: 'id', 'sentiment', and 'explanation'. ",
    "Do not include markdown formatting, markdown code blocks (like ```json), ",
    "or any conversational filler.\n\n",
    "Data:\n", batch_json_string
  )

  req <- request(url) %>%
    req_body_json(list(
      contents = list(
        list(parts = list(list(text = prompt)))
      ),
      generationConfig = list(
        temperature = 0.0   # Keep output deterministic for classification
      )
    )) %>%
    req_retry(max_tries = 3)  # Automatically retries on HTTP 503

  # Send request; catch network or HTTP errors
  api_response <- tryCatch({
    resp <- req_perform(req)
    resp_body_json(resp)
  }, error = function(e) {
    message(sprintf("API error in batch %d: %s", i, e$message))
    return(NULL)
  })

  # --- Handle the response ---
  batch_success <- FALSE

  if (!is.null(api_response)) {
    raw_text <- api_response$candidates[[1]]$content$parts[[1]]$text

    parsed_response <- tryCatch({
      clean_text <- gsub("```json|```", "", raw_text)
      clean_text <- trimws(clean_text)
      jsonlite::fromJSON(clean_text)
    }, error = function(e) {
      message(sprintf("JSON parse error in batch %d — raw response was malformed.", i))
      return(NULL)
    })

    if (!is.null(parsed_response)) {
      results_list[[as.character(i)]] <- as_tibble(parsed_response)
      batch_success <- TRUE
    }
  }

  # FIX (Issue 3): If the batch failed at any point, log its ID immediately
  # to a file so you have a record without having to scan for NAs afterwards.
  if (!batch_success) {
    failed_batches <- c(failed_batches, i)
    write(
      sprintf("FAILED batch %d at %s", i, Sys.time()),
      file = "Data/failed_batches.log",
      append = TRUE
    )
  }

  # Progress message every 10 batches
  if (i %% 10 == 0) {
    message(sprintf(
      "Processed batch %d of %d (%.1f%%) — %d failed so far",
      i, total_batches, (i / total_batches) * 100, length(failed_batches)
    ))
  }

  # FIX (Issue 4): Save a checkpoint every 15 batches instead of every 50.
  # With ~1,480 batches, saving every 50 means up to 49 batches of lost work
  # after a crash. Every 15 limits that loss to at most 14 batches.
  if (i %% 15 == 0) {
    saveRDS(results_list, "Data/partial_results.rds")
    message(sprintf("Checkpoint saved at batch %d.", i))
  }

  # FIX (Issue 5): Only sleep after a *successful* call.
  # Sleeping after a failure wastes time and gains nothing.
  if (batch_success) Sys.sleep(0.5)
}

# Final checkpoint save
saveRDS(results_list, "Data/partial_results.rds")

# Summary of failures
if (length(failed_batches) > 0) {
  message(sprintf(
    "Done. %d batch(es) failed: %s. See failed_batches.log for details.",
    length(failed_batches), paste(failed_batches, collapse = ", ")
  ))
} else {
  message("Done. All batches completed successfully.")
}
```

#### 5.4 Combine and save results

```{r save-results}
# Combine all successful batch results into one data frame
final_sentiments <- bind_rows(results_list) |> 
  distinct(id, .keep_all = TRUE) #Avoid having duplicates in the results

# Join back to the original data (tweets with no result will show NA)
final_data <- my_data |>  left_join(final_sentiments, by = "id")

write.csv(final_data, "Data/Gemini_results.csv", row.names = FALSE)

message(sprintf(
  "Saved %d rows. Rows with missing sentiment (NA): %d",
  nrow(final_data),
  sum(is.na(final_data$sentiment))
))
```

The output has produced the same number of rows but there are 1355 rows with missing data (NA). This is because some tweets included special characters (like quotation marks), which caused Gemini to generate invalid JSON that could not be read or processed.

#### 5.5 Re-run the code on missing batches

```{r retry-failed-batches}
# ── RETRY CHUNK ──────────────────────────────────────────────────────────────
# Runs only the 66 batches that failed in the main loop.
# Safe to re-run multiple times — skips any batch already in results_list.
# ─────────────────────────────────────────────────────────────────────────────

# Load the checkpoint so we don't lose the 1,414 batches that succeeded
results_list <- readRDS("Data/partial_results.rds")

# The 66 failed batch IDs from the log
failed_batches <- c(22, 27, 34, 40, 90, 103, 107, 118, 119, 143, 147, 183,
                    190, 210, 211, 215, 241, 270, 301, 306, 310, 378, 409,
                    416, 488, 537, 679, 709, 724, 744, 747, 760, 766, 809,
                    835, 866, 881, 887, 893, 915, 924, 925, 937, 963, 975,
                    986, 993, 1000, 1022, 1064, 1071, 1082, 1147, 1179, 1192,
                    1202, 1251, 1296, 1301, 1317, 1355, 1365, 1371, 1376,
                    1436, 1454)

still_failed <- c()

for (i in failed_batches) {

  # Skip if already recovered in a previous retry run
  if (as.character(i) %in% names(results_list)) {
    message(sprintf("Batch %d already recovered — skipping.", i))
    next
  }

  current_batch <- my_data %>% filter(batch_id == i)

  # Extra aggressive quote cleaning for problem batches:
  # Replace ALL types of quotes and apostrophes with a simple apostrophe
  current_batch <- current_batch %>%
    mutate(text_clean = str_replace_all(text_clean, '[\u201C\u201D\u201E\u201F\u0022]', "'"),
           text_clean = str_replace_all(text_clean, '[\u2018\u2019\u201A\u201B]', "'"),
           text_clean = str_replace_all(text_clean, '\\\\', ' '))  # remove backslashes

  batch_json_string <- jsonlite::toJSON(
    current_batch %>% select(id, text_clean),
    auto_unbox = TRUE
  )

  prompt <- paste0(
    "You are a strict data processing tool. Do sentiment analysis on the ",
    "text data provided in JSON format below following these instructions:\n\n",
    "1. Sentiment Classification: In English, classify each tweet exactly as ",
    "'positive', 'negative', or 'neutral' (choose only one label per tweet).\n",
    "2. Explanation: Provide a concise and specific explanation for how you ",
    "decided on that label.\n\n",
    "CRITICAL: Return ONLY a valid JSON array of objects containing exactly ",
    "three keys: 'id', 'sentiment', and 'explanation'. ",
    "Do not include markdown formatting, markdown code blocks (like ```json), ",
    "or any conversational filler.\n\n",
    "Data:\n", batch_json_string
  )

  req <- request(url) %>%
    req_body_json(list(
      contents = list(
        list(parts = list(list(text = prompt)))
      ),
      generationConfig = list(temperature = 0.0)
    )) %>%
    req_retry(max_tries = 3)

  api_response <- tryCatch({
    resp <- req_perform(req)
    resp_body_json(resp)
  }, error = function(e) {
    message(sprintf("API error in batch %d: %s", i, e$message))
    return(NULL)
  })

  batch_success <- FALSE

  if (!is.null(api_response)) {
    raw_text <- api_response$candidates[[1]]$content$parts[[1]]$text

    parsed_response <- tryCatch({
      clean_text <- gsub("```json|```", "", raw_text)
      clean_text <- trimws(clean_text)
      jsonlite::fromJSON(clean_text)
    }, error = function(e) {
      message(sprintf("Still failing — batch %d malformed again.", i))
      return(NULL)
    })

    if (!is.null(parsed_response)) {
      results_list[[as.character(i)]] <- as_tibble(parsed_response)
      batch_success <- TRUE
      message(sprintf("✓ Recovered batch %d", i))
    }
  }

  if (!batch_success) {
    still_failed <- c(still_failed, i)
  }

  if (batch_success) Sys.sleep(0.5)
}

# Save updated results
saveRDS(results_list, "Data/partial_results.rds")

# Summary
message(sprintf("\nRetry complete. Recovered %d of %d failed batches.",
                length(failed_batches) - length(still_failed), length(failed_batches)))
if (length(still_failed) > 0) {
  message(sprintf("Still failing after retry: %s", paste(still_failed, collapse = ", ")))
} else {
  message("All previously failed batches recovered successfully!")
}
```

After retrying the code I have recovered 4 of 66 failed batches only. Those batches have tweets so heavily corrupted with special characters that even the aggressive cleaning is not enough.

##### Save results

```{r rebuild-final-csv}

results_list <- readRDS("Data/partial_results.rds")

# Rebuild the final CSV with the recovered batches included
final_sentiments <- bind_rows(results_list)

final_data <- my_data %>%
  left_join(final_sentiments, by = "id") %>%
  distinct(id, .keep_all = TRUE)  # remove duplicate rows

write.csv(final_data, "Data/Gemini_results.csv", row.names = FALSE)

message(sprintf(
  "Saved %d rows. Rows with missing sentiment (NA): %d",
  nrow(final_data),
  sum(is.na(final_data$sentiment))
))
```

### 6.Standarisation

The standarisation process requieres that the data is numeric. Since there are NA's in the Gemini results, numeric matrices handle NA seamlessly for calculating Krippendorff's Alpha

#### 6.1 NRC

```{r}
nrc_results <- read.csv("Data/NRC_results.csv", colClasses = c(id = "character"))
```

##### Function to select one single label

```{r}
choose_label <- function(data) {
  if(data$negative > data$positive) {
    label <- "negative"
  } else if (data$positive > data$negative) {
    label <- "positive"
  } else {
    label <- "neutral" # create an extra label, if negative = positive then it is neutral
  }
  return(label)
}
```

##### Run the function on the database

```{r}
nrc_results$nrc_lexicon <- sapply(1:nrow(nrc_results), function(i) choose_label(nrc_results[i, ]))
```

##### Save the new dataframe 

Substitute the old csv with new one, where only a new column is added ("nrc_lexicon")

```{r}
write.csv(nrc_results, "Data/NRC_results.csv", row.names = FALSE)
```

#### 6.2 RoBERTuito

```{r}
robertuito_results <- read.csv("Data/RoBERTuito_results.csv", colClasses = c(id = "character"))
```

##### Change labels

The RoBERTuito model outputs three labels: POS, NEG and NEU. In order to make a comparison between models the labels should be standarised.

```{r}
robertuito_results <- robertuito_results |> 
  mutate(robertuito_transformer = case_when(
    label == "POS" ~ "positive",
    label == "NEG" ~ "negative",
    label == "NEU" ~ "neutral"
  ))
```

##### save the results

```{r}
write.csv(robertuito_results, "Data/RoBERTuito_results.csv", row.names = FALSE)
```

#### 6.3 Gemini

Load the Gemini results

```{r read Gemini results}
gemini_results <- read.csv("Data/Gemini_results.csv", colClasses = c(id = "character"))
```

```{r add gemini llm column}
gemini_results <- gemini_results |> select(-text_clean.y) |> 
  mutate(gemini_llm = sentiment)
```

```{r}
write.csv(gemini_results, "Data/Gemini_results.csv", row.names = FALSE)
```

### 7. Join dataframes 

The data will be joined by id, producing a new data frame with the results for each model.

```{r combine results}
combined_results <- list(
  nrc_results |> select(id, political_party, tweet, nrc_lexicon),
  robertuito_results |> select(id, robertuito_transformer),
  gemini_results |> select(id, gemini_llm)) |> 
  reduce(full_join, by = "id")
```

```{r}
write.csv(combined_results, "Data/models_combined_results.csv", row.names = FALSE)
```

### 8. Sentiment distributions and Krippendorff's Alpha 
#### 8.1 Libraries

```{r libraries}
library(irr) # library Krippendorff's Alpha
library(gt) # to create tables
library(purrr)
```

#### 8.2 Load data and shared factor definitions

```{r load combined results}
combined_results <- read.csv("Data/models_combined_results.csv", colClasses = c(id = "character"))
```

```{r models and party labels}
# ── Model recoding ──────────────────────────────────────────────────────────
model_labels <- c(
  "nrc_lexicon"            = "NRC",
  "robertuito_transformer" = "RoBERTuito",
  "gemini_llm"             = "Gemini"
)
model_levels <- c("NRC", "Gemini", "RoBERTuito")   # left-to-right display order

# ── Party recoding (full names, used in most outputs) ────────────────────────
party_labels_full <- c(
  "esquerra republicana" = "Esquerra Republicana de Catalunya (ERC)",
  "podemos"              = "Podemos",
  "psoe"                 = "PSOE",
  "ciudadanos"           = "Ciudadanos",
  "partido popular"      = "Partido Popular (PP)",
  "vox"                  = "Vox"
)
# Ideological left-to-right ordering (full names)
party_levels_full <- c(
  "Esquerra Republicana de Catalunya (ERC)", "Podemos", "PSOE",
  "Ciudadanos", "Partido Popular (PP)", "Vox"
)

# ── Party recoding (abbreviated names, used in the heatmap) ─────────────────
party_labels_abbr <- c(
  "esquerra republicana" = "ERC",
  "podemos"              = "Podemos",
  "psoe"                 = "PSOE",
  "ciudadanos"           = "Ciudadanos",
  "partido popular"      = "PP",
  "vox"                  = "Vox"
)
party_levels_abbr <- c("ERC", "Podemos", "PSOE", "Ciudadanos", "PP", "Vox")

# ── Sentiment levels ─────────────────────────────────────────────────────────
sentiment_levels <- c("negative", "neutral", "positive")

# ── Shared colour palette ────────────────────────────────────────────────────
sentiment_colours <- c(
  "positive" = "#2ecc71",
  "negative" = "#e74c3c",
  "neutral"  = "#3498db"
)

# ── Helper: convert a character model column to ordered numeric (for kripp.alpha)
to_numeric_sentiment <- function(x) {
  as.numeric(factor(x, levels = sentiment_levels))
}

# ── Helper: compute Krippendorff's alpha for two model columns ───────────────
calc_kripp_pair <- function(data, model1, model2) {
  pair_matrix <- data |>
    select(all_of(c(model1, model2))) |>
    mutate(across(everything(), to_numeric_sentiment)) |>
    t()
  kripp.alpha(pair_matrix, method = "nominal")
}

# ── Model pairs used in both global and party-level analyses ─────────────────
pairs <- list(
  c("nrc_lexicon", "robertuito_transformer"),
  c("nrc_lexicon", "gemini_llm"),
  c("robertuito_transformer", "gemini_llm")
)

```

#### 8.3 Overall sentiment distribution

##### Visualisation: overall sentiment distribution by model

```{r global-sentiment-distribution}
# CHANGED: uses model_labels, model_levels, sentiment_colours defined above
combined_results |>
  pivot_longer(cols = c(nrc_lexicon, robertuito_transformer, gemini_llm),
               names_to  = "model",
               values_to = "sentiment") |>
  mutate(
    model     = factor(recode(model, !!!model_labels), levels = model_levels),
    sentiment = factor(sentiment, levels = c(sentiment_levels, NA))
  ) |>
  count(model, sentiment) |>
  group_by(model) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup() |>
  complete(model, sentiment, fill = list(n = 0, pct = 0)) |>
  ggplot(aes(x = sentiment, y = pct, fill = sentiment)) +
  geom_col() +
  geom_text(aes(label = paste0(pct, "%")),
            vjust = -0.4, size = 3, color = "gray30") +
  facet_wrap(~ model) +
  scale_fill_manual(values = c(sentiment_colours, "NA" = "gray")) +
  labs(title = "Sentiment distribution by model",
       x = "Sentiment", y = "Percentage") +
  scale_y_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, by = 20),
                     labels = paste0(seq(0, 60, by = 20), "%")) +
  theme_minimal() +
  theme(panel.grid      = element_blank(),
        strip.text      = element_text(size = 12),
        axis.text.x     = element_blank(),
        axis.ticks.x    = element_blank(),
        axis.ticks.y    = element_line(color = "black", linewidth = 0.5),
        axis.ticks.length = unit(0.2, "cm"),
        strip.background = element_blank())
```

```{r}
ggsave("Figures_and_tables/Models_sentiment_distribution.png", width = 8, height = 5, dpi = 300)
```

#### 8.4 Party-specific and political sentiment distribution

##### Visualisation: sentiment distribution by model and party

```{r party-sentiment-distribution,  fig.height= 7, fig.width= 6}
plot_by_model_party <- combined_results |>
  pivot_longer(cols      = c(nrc_lexicon, robertuito_transformer, gemini_llm),
               names_to  = "model",
               values_to = "sentiment") |>
  filter(!is.na(sentiment)) |>
  mutate(
    political_party = factor(recode(political_party, !!!party_labels_full),
                             levels = party_levels_full),
    model           = factor(recode(model, !!!model_labels),
                             levels = model_levels),
    sentiment       = factor(sentiment, levels = sentiment_levels)
  ) |>
  count(political_party, model, sentiment) |>
  group_by(political_party, model) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup() |>
  ggplot(aes(x = sentiment, y = pct, fill = sentiment)) +
  geom_col(width = 1.0) +
  facet_grid(political_party ~ model) +
  scale_fill_manual(values = sentiment_colours) +
  scale_y_continuous(limits = c(0, 75), expand = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(expand = expansion(add = 2)) + 
  labs(title = "Sentiment distribution by model and party",
       x = "Sentiment distribution", y = "Percentage") +
  theme_minimal() +
  theme(legend.position   = "bottom",
        panel.grid        = element_blank(),
        strip.text.x      = element_text(size = 11),
        strip.text.y      = element_text(size = 10, angle = 0, hjust = 0),
        strip.background  = element_blank(),
        axis.text.x       = element_blank(),
        axis.ticks.x      = element_blank(),
        panel.border      = element_blank(),
        panel.spacing.y   = unit(0.2, "lines"),              
        axis.text.y       = element_text(size = 6.5),
        panel.grid.major.y = element_line(color = "grey90"))
plot_by_model_party
```

```{r party-sentiment-distribution,  fig.height= 7, fig.width= 6}
library(patchwork)
library(grid)

arrow_plot <- ggplot() +
  # Political Spectrum label in middle
  annotate("text", x = 0.5, y = 0.5, label = "Political\nSpectrum", 
           size = 3, fontface = "bold", hjust = 0.5) +
  # top arrow (Left-wing)
  annotate("segment", x = 0.5, xend = 0.5, y = 0.55, yend = 1,
           arrow = grid::arrow(length = unit(0.2, "cm")), color = "black") +
  annotate("text", x = 0.7, y = 0.77, label = "Left", 
           size = 3.5, angle = 270, hjust = 0.5) +
  # bottom arrow (Right-wing)
  annotate("segment", x = 0.5, xend = 0.5, y = 0.45, yend = 0,
           arrow = grid::arrow(length = unit(0.2, "cm")), color = "black") +
  annotate("text", x = 0.7, y = 0.23, label = "Right", 
           size = 3.5, angle = 270, hjust = 0.5) +
  theme_void() +
  xlim(0, 1) + ylim(0, 1)


arrow_plot
```


```{r party-sentiment-distribution,  fig.height= 7, fig.width= 7}
plot_by_model_party + arrow_plot + plot_layout(widths = c(10, 3))
```

```{r}
ggsave("Figures_and_tables/Party_sentiment_distribution.png", width = 7, height = 7, dpi = 300)
```

##### Standard deviation of sentiment percentages across parties

```{r sd-by-party}
sd_summary <- combined_results |>
  pivot_longer(cols      = c(nrc_lexicon, robertuito_transformer, gemini_llm),
               names_to  = "model",
               values_to = "sentiment") |>
  filter(!is.na(sentiment)) |>
  mutate(model = factor(recode(model, !!!model_labels), levels = model_levels)) |>
  count(political_party, model, sentiment) |>
  group_by(political_party, model) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  group_by(model) |>
  summarise(sd = round(sd(pct), 2))

sd_summary_table <- sd_summary |>
  gt() |>
  tab_header(
    title    = "Sentiment Variance Across Political Parties",
    subtitle = "Standard deviation of sentiment percentages by model"
  ) |>
  cols_label(model = "Model", sd = "SD") |>
  tab_options(table.width = pct(50))

sd_summary_table
```

```{r}
gtsave(sd_summary_table, "Figures_and_tables/sd_party_table.png")
```

Interpretation: the standard deviation indicates how much sentiment percentages
vary across parties within each model.

- Low SD (NRC) → the model assigns roughly the same sentiment distribution
  regardless of party — it is not sensitive to party identity.
- High SD (RoBERTuito / Gemini) → sentiment percentages vary considerably by
  party — the model responds differently to different parties' language.

#### 8.5 Inter-model reliability and Agreement Analysis

##### Krippendorff's Alpha (all three models)

```{r global-kripp-alpha}
# CHANGED: uses to_numeric_sentiment() helper defined above instead of
# repeating the inline factor() call
kripp_matrix <- combined_results |>
  select(nrc_lexicon, robertuito_transformer, gemini_llm) |>
  mutate(across(everything(), to_numeric_sentiment)) |>
  t()

krippendorff <- kripp.alpha(kripp_matrix, method = "nominal")
print(krippendorff)
```

```{r global-kripp-table}
# CHANGED: references model_labels for display names instead of hard-coding
kripp_table <- tibble(
  models = paste(model_labels, collapse = ", "),
  alpha  = krippendorff$value
) |>
  gt() |>
  fmt_number(columns = alpha, decimals = 3) |>
  tab_header(title = "Global Krippendorff's Alpha") |>
  cols_label(models = "Models", alpha = "α") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels())

kripp_table
```

```{r}
gtsave(kripp_table, "Figures_and_tables/kripp_table.png")
```
Interpretation: the Krippendorff’s Alpharanges from -1 to 1, where:
- 1 indicates perfect agreement between the models
- 0 indicates agreement no better than chance
- -1 indicates systematic disagreement, where models tend to assign opposing labels more often than would be expected randomly. 

##### Krippendorff's Alpha by pairs of models

The single alpha above summarises all three models. Computing pairwise alphas
reveals which specific pair diverges most.

```{r global-kripp-pairs}
# CHANGED: uses model_labels for recode; no repetition of label strings
kripp_pairs <- map_dfr(pairs, function(pair) {
  result <- calc_kripp_pair(combined_results, pair[1], pair[2])
  tibble(model_1 = pair[1], model_2 = pair[2], alpha = result$value)
})

kripp_pairs_table <- kripp_pairs |>
  mutate(
    model_1 = recode(model_1, !!!model_labels),   # CHANGED: !!! splices model_labels
    model_2 = recode(model_2, !!!model_labels)
  ) |>
  gt() |>
  tab_header(
    title    = "Krippendorff's Alpha by Model Pair",
    subtitle = "Nominal scale — pairwise inter-rater agreement"
  ) |>
  cols_label(model_1 = "Model 1", model_2 = "Model 2", alpha = "α") |>
  fmt_number(columns = alpha, decimals = 3) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_spanner(label = "Model Pair", columns = c(model_1, model_2))

kripp_pairs_table
```

```{r}
gtsave(kripp_pairs_table, "Figures_and_tables/kripp_pairs_table.png")
```

#### 8.4 Party-level Agreement Analysis

##### Krippendorff's Alpha by party

```{r kripp-by-party}
kripp_by_party <- combined_results |>
  group_by(political_party) |>
  group_map(.keep = TRUE, ~ {
    mat <- .x |>
      select(nrc_lexicon, robertuito_transformer, gemini_llm) |>
      mutate(across(everything(), to_numeric_sentiment)) |>
      t()
    result <- kripp.alpha(mat, method = "nominal")
    tibble(
      political_party = .x$political_party[1],
      subjects        = result$subjects,
      raters          = result$raters,
      alpha           = round(result$value, 3)
    )
  }) |>
  bind_rows() |>
  mutate(party = factor(recode(political_party, !!!party_labels_full),
                        levels = party_levels_full))   # CHANGED: splices shared vectors

party_kripp_table <- kripp_by_party |>
  select(party, subjects, raters, alpha) |>
  arrange(party) |>
  gt() |>
  cols_label(party    = "Party",
             subjects = "Subjects",
             raters   = "Raters",
             alpha    = "Krippendorff's α") |>
  tab_header(title = "Inter-model Agreement by Political Party") |>
  fmt_number(columns = alpha, decimals = 3) |>
  tab_style(style     = cell_text(weight = "bold"),
            locations = cells_column_labels())

party_kripp_table
```

```{r}
gtsave(party_kripp_table, "Figures_and_tables/krippendorff_party_table.png")
```

---

##### Krippendorff's Alpha by party and model pair (line graph)

```{r kripp-pairs-by-party}
# Data: pairwise Krippendorff's Alpha per party, ordered left-to-right ideologically
kripp_pairs_party <- combined_results |>
  group_by(political_party) |>
  group_map(.keep = TRUE, ~ {
    map_dfr(pairs, function(pair) {
      result <- calc_kripp_pair(.x, pair[1], pair[2])
      tibble(
        political_party = .x$political_party[1],
        model_1         = pair[1],
        model_2         = pair[2],
        alpha           = result$value
      )
    })
  }) |>
  bind_rows() |>
  mutate(
    pair = paste(recode(model_1, !!!model_labels),
                 "vs",
                 recode(model_2, !!!model_labels)),
    political_party = factor(recode(political_party, !!!party_labels_abbr),
                             levels = party_levels_abbr)  # ideological left-to-right order
  )

# Add global (all-model) alpha per party as a fourth line
kripp_global_party <- combined_results |>
  group_by(political_party) |>
  group_map(.keep = TRUE, ~ {
    mat <- .x |>
      select(nrc_lexicon, robertuito_transformer, gemini_llm) |>
      mutate(across(everything(), to_numeric_sentiment)) |>
      t()
    result <- kripp.alpha(mat, method = "nominal")
    tibble(
      political_party = .x$political_party[1],
      pair            = "All Models Combined",
      alpha           = result$value
    )
  }) |>
  bind_rows() |>
  mutate(political_party = factor(recode(political_party, !!!party_labels_abbr),
                                  levels = party_levels_abbr))

# Combine pairwise and global lines
kripp_lines_data <- bind_rows(
  kripp_pairs_party |> select(political_party, pair, alpha),
  kripp_global_party
) |>
  mutate(pair = factor(pair, levels = c(
    "NRC vs RoBERTuito",
    "NRC vs Gemini",
    "RoBERTuito vs Gemini",
    "All Models Combined"
  )))

# Line colours: NRC pairs muted, RoBERTuito-Gemini highlighted, global dashed
pair_colours <- c(
  "NRC vs RoBERTuito"    = "#e74c3c",
  "NRC vs Gemini"        = "#e67e22",
  "RoBERTuito vs Gemini" = "#2ecc71",
  "All Models Combined"  = "#2c3e50"
)

pair_linetypes <- c(
  "NRC vs RoBERTuito"    = "solid",
  "NRC vs Gemini"        = "solid",
  "RoBERTuito vs Gemini" = "solid",
  "All Models Combined"  = "dashed"
)

kripp_linegraph_party_pairs <- ggplot(
  kripp_lines_data,
  aes(x = political_party, y = alpha,
      colour = pair, linetype = pair, group = pair)
) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.4) +  # zero reference
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = pair_colours, name = "Model Pair") +
  scale_linetype_manual(values = pair_linetypes, name = "Model Pair") +
  scale_y_continuous(
    limits = c(-0.2, 0.6),
    breaks = seq(-0.2, 0.6, by = 0.1)
  ) +
  labs(
    title   = "Krippendorff's Alpha by Model Pair and Political Party",
    subtitle = "Parties ordered left to right on the political spectrum",
    x       = "Political spectrum (left → right)",
    y       = "Krippendorff's Alpha (α)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90")
  )

kripp_linegraph_party_pairs
```

```{r}
ggsave("Figures_and_tables/kripp_linegraph_party_pairs.png", width = 9, height = 5.5, dpi = 300)
```


### 9. Create sub-dataframes 

```{r load combined results}
combined_results <- read.csv("Data/models_combined_results.csv", colClasses = c(id = "character"))
```

```{r}
nrc_results <- read.csv("Data/NRC_results.csv", colClasses = c(id = "character"))
```

### 9.1 Subdataframe of agreement results

The following data frame is a subset of tweets where all of the models produced the same output for the same tweet id.

```{r}
perfect_agreement_df <- combined_results |> 
  filter(
    !is.na(gemini_llm),
    nrc_lexicon == robertuito_transformer & nrc_lexicon == gemini_llm) |> 
  # Join with the dataset that contains the tweet text and political party
  left_join(
    nrc_results |> select(id, political_party, tweet), by = "id") |> 
  relocate(political_party, tweet, .after = id) #Move political_party and tweet right after the 'id' column

perfect_agreement_df
```

```{r}
write.csv(perfect_agreement_df, "agreement_and_disagreement_df/Perfect_agreement_df.csv", row.names = FALSE)
```

### 9.1 Subdataframe of disagreement results

```{r}
disagreement_df <- combined_results |> 
  # 1. Clean out any rows with NA values first
  filter(!is.na(gemini_llm)) |> 
  # 2. Drop any row that is found in your perfect agreement dataframe
  anti_join(perfect_agreement_df, by = "id") |> 
  left_join(
    nrc_results |> select(id, political_party, tweet), by = "id") |> 
  relocate(political_party, tweet, .after = id) #Move political_party and tweet right after the 'id' column

disagreement_df
```

```{r}
write.csv(disagreement_df, "agreement_and_disagreement_df/Disagreement_df.csv", row.names = FALSE)
```

#### 9.3 Subdataframe of NA's in Gemini

```{r}
missing_gemini <- combined_results |> 
  filter(is.na(gemini_llm)) |> 
  relocate(political_party, tweet, .after = id)

missing_gemini
```

```{r}
write.csv(missing_gemini, "agreement_and_disagreement_df/Gemini_missing_df.csv", row.names = FALSE)
```

