---
title: "Sentiment analysis twitter"
author: "Miriam Lim Leal"
date: "2026-03-26"
output: html_document
---

## Sentiment analysis Twitter

```{r}
library(tidyverse)
library(tidytext)
library(textdata)
library(emoji)
library(readxl)
library(stringi)
library(syuzhet)
library(udpipe)
```

### 1. Load data

```{r}
twitter_data <- read.csv("tweets_political_parties_es.csv", colClasses = c(id = "character"))
```

### 2. Clean tweet text

#### General pipeline

Remove URLs, mentions and normalise whitespace

```{r}
clean_general <- function(text) {
  
  # Remove URLs
  text <- str_remove_all(text, "https?://\\S+")
  
  # Remove mentions
  text <- str_remove_all(text, "@\\w+")
  
  # Normalize whitespace
  text <- str_squish(text)
  
  return(text)
}
```

```{r}
twitter_transformer <- twitter_data |> 
  mutate(text_clean = map_chr(tweet, clean_general))
```

### 3. Lexicon-based model

#### 3.1 Emoji mapping in Spanish

##### Load the Spanish emoji mapping (Emoji-SP).

```{r}
emoji_map <- read_excel("emoji_mapping_es.xlsx") |> 
  mutate(definition_lower = str_to_lower(definition)) |> 
  select(emoji, definition_lower, val)
```

##### Function replace emojis with text using Emoji-SP

```{r}
# This replaces the emoji symbol with the Spanish definition from Emoji-SP file
replace_emoji_with_def <- function(text, lexicon) {
  # We use stri_replace_all_fixed to swap emoji symbols for definitions
  # We add spaces and a prefix 'EMOJI_' to help the model distinguish it
  new_text <- stri_replace_all_fixed(
    text, 
    lexicon$emoji, 
    paste0(" ", lexicon$definition_lower, " "), 
    vectorize_all = FALSE
  )
  return(new_text)
}
```

##### Create a new column with emoji replaced text called emoji_text

```{r}
twitter_lexicon <- twitter_transformer %>%
  mutate(emoji_text = sapply(text_clean, replace_emoji_with_def, lexicon = emoji_map))
```

#### 3.2 Lexicon-specific cleaning pipeline

clean data (specifically column emoji_text): drop hashtags symbols, remove punctuation and make everything lowercase.

```{r}
clean_lexicon <- function(text) {
  
  # Keep hashtag words, drop the symbol
  text <- str_replace_all(text, "#(\\w+)", "\\1")
  
  # Remove all punctuation
  text <- str_remove_all(text, "[!\"#$%&'()*+,\\-./:;<=>?@\\[\\\\\\]^_`{|}~¡¿]")
  
  # Lowercase for lexicon matching
  text <- str_to_lower(text)
  
  # Normalize whitespace again after removals
  text <- str_squish(text)
  
  return(text)
}
```

```{r}
twitter_lexicon <- twitter_data |> 
  mutate(
    # Step 1: shared cleaning (URLs, mentions, whitespace)
    text_clean = map_chr(tweet, clean_general),
    
    # Step 2: replace emojis with Spanish definitions (applied to text_clean)
    emoji_text        = map_chr(text_clean, replace_emoji_with_def, lexicon = emoji_map),
    
    # Step 3: lexicon cleaning (hashtags, punctuation, lowercase)
    clean_lexicon_text = map_chr(emoji_text, clean_lexicon)
  ) |>
  select(-emoji_text)    
```

Save this as a csv

```{r}
# write.csv(twitter_lexicon, "twitter_lexicon_and_transformer_clean.csv", row.names = FALSE)
```

#### 3.3. Lexicon model: NRC

Lemmatize the text function

```{r}
# Function to lemmatize Spanish text
lemmatize_spanish <- function(text, model) {
  # If text is empty, return empty
  if (nchar(text) == 0) return("")
  
  # Annotate the text
  annotation <- udpipe_annotate(model, x = text)
  df <- as.data.frame(annotation)
  
  # Paste the 'lemma' column back together into a single string
  lemmatized_text <- paste(df$lemma, collapse = " ")
  return(lemmatized_text)
}
```

Download the Spanish model

```{r}
# udpipe_download_model(language = "spanish")  # only download once
ud_model <- udpipe_load_model("spanish-gsd-ud-2.5-191206.udpipe")
```

REVISE LEMMATISATION ----------------------------------------------------

Apply lemmatisation to the text (clean_lexicon_text)

```{r}
# twitter_lexicon <- twitter_lexicon |> 
#  mutate(lemmatized_text = map_chr(clean_lexicon_text, lemmatize_spanish, model = ud_model))
```

After it has been run once, save to use for other models

```{r}
#write.csv(twitter_lexicon, "twitter_lexicon_lemmatized.csv", row.names = FALSE)
```

END OF LEMMATISATION -----------------------------------------------------

##### Run NRC

```{r}
nrc_results <- get_nrc_sentiment(twitter_lexicon$clean_lexicon_text, language = "spanish")
```

```{r}
# Bind the id and tweet columns from twitter_lexicon to the NRC results
nrc_results <- bind_cols(
  twitter_lexicon |> select(id, political_party, tweet),
  nrc_results
)
```

##### Save the results of NRC

```{r}
write.csv(nrc_results, "NRC_results.csv", row.names = FALSE)
```

### 4. Trasformer-based model: RoBERTuito

#### 4.1 Open clean data

```{r}
clean_data <- read.csv("twitter_lexicon_and_transformer_clean.csv", colClasses = c(id = "character"))
```

#### 4.2 Install prerequisites

```{r}
#install.packages("reticulate")
library(reticulate)

# Install Python dependencies:
# py_install(c("transformers", "torch", "pysentimiento"), pip = TRUE)
```

```{r}
pysentimiento <- import("pysentimiento")
```

#### 4.3 Create analyzer for sentiment

```{r}
analyzer <- pysentimiento$create_analyzer(task = "sentiment", lang = "es")
```

#### 4.4 Apply to data

```{r}
results <- lapply(clean_data$text_clean, function(x) {
  pred <- analyzer$predict(x)
  data.frame(
    label = pred$output,
    POS   = pred$probas[["POS"]],
    NEG   = pred$probas[["NEG"]],
    NEU   = pred$probas[["NEU"]]
  )
})
```

#### 4.5 Create 1 single dataframe:

```{r}
results_df <- do.call(rbind, results)
```

#### 4.6 join results to original dataframe

Only include id, political_party and tweet from "clean_data" and bind with the results from RoBERTuito

```{r}
robertuito_results <- cbind(clean_data[, c("id", "political_party", "tweet")], results_df)
```

#### 4.7 Save results

```{r}
write.csv(robertuito_results, "RoBERTuito_results.csv", row.names = FALSE)
```



```{r}
nrc_results <- read.csv("NRC_results.csv", colClasses = c(id = "character"))
```

```{r}

## REVISEEEEEEEEEEE
nrc_results <- nrc_results |>
  mutate(political_party = clean_data$political_party) |>
  select(id, political_party, everything())

write.csv(nrc_results, "NRC_results.csv", row.names = FALSE)
```

### 5. LLM model: Gemini 

Gemini 3.1 Flash Lite model for the analysis

- Batching: Group data into batches of 10 to 20 rows and ask Gemini to return a structured JSON list of results.
- Structured Output: Force Gemini to return only valid JSON
- Rate Limit Buffering: Implement a small Sys.sleep() time delay between batches.

#### 5.1 Setup

```{r libraries}
library(httr2)
library(tidyverse)
library(jsonlite)

```

#### 5.2 Prepare data 

Read data and select only the columns needed

```{r prepare-data}
my_data <- read.csv("twitter_lexicon_and_transformer_clean.csv", colClasses = c(id = "character"))

my_data <- my_data |> select(id, political_party, text_clean)

# Normalise quotes to prevent JSON breaking
my_data <- my_data %>%
  mutate(text_clean = str_replace_all(text_clean, '[\u201C\u201D\u201E\u201F]', '"'),
         text_clean = str_replace_all(text_clean, '"', "'"))

# Define batch size (e.g., 20 rows per API call)
batch_size <- 20
my_data <- my_data %>% 
  mutate(batch_id = ceiling(row_number() / batch_size))
```

#### 5.3 Analysis Loop

```{r gemini-loop}

API_KEY <- Sys.getenv("GEMINI_KEY") #API key is saved in .Renviron file as: GEMINI_API_KEY=your_key, call usethis::edit_r_environ() to open that file and save it.

if (API_KEY == "") stop("GEMINI_KEY not found. Check your .Renviron file.")

# Select model: gemini-2.5-flash-lite, which is designed for simple classification tasks (gemini-2.5-flash is much more expensive) 
url <- paste0(
  "https://generativelanguage.googleapis.com/v1beta/models/",
  "gemini-2.5-flash-lite:generateContent?key=", API_KEY
)

# Initialise results storage and failure log
results_list  <- list()
failed_batches <- c()   # track failed batch IDs in memory
total_batches  <- max(my_data$batch_id)

for (i in 1:total_batches) {

  # Skip this batch if it was already processed.
  # This prevents silent duplicates if the loop is re-run after a crash.
  if (as.character(i) %in% names(results_list)) {
    message(sprintf("Batch %d already processed — skipping.", i))
    next
  }

  current_batch <- my_data %>% filter(batch_id == i)

  # jsonlite::toJSON handles all quote escaping internally — do not pre-escape.
  batch_json_string <- jsonlite::toJSON(
    current_batch %>% select(id, text_clean),
    auto_unbox = TRUE
  )

  prompt <- paste0(
    "You are a strict data processing tool. Do sentiment analysis on the ",
    "text data provided in JSON format below following these instructions:\n\n",
    "1. Sentiment Classification: In English, classify each tweet exactly as ",
    "'positive', 'negative', or 'neutral' (choose only one label per tweet).\n",
    "2. Explanation: Provide a concise and specific explanation for how you ",
    "decided on that label.\n\n",
    "CRITICAL: Return ONLY a valid JSON array of objects containing exactly ",
    "three keys: 'id', 'sentiment', and 'explanation'. ",
    "Do not include markdown formatting, markdown code blocks (like ```json), ",
    "or any conversational filler.\n\n",
    "Data:\n", batch_json_string
  )

  req <- request(url) %>%
    req_body_json(list(
      contents = list(
        list(parts = list(list(text = prompt)))
      ),
      generationConfig = list(
        temperature = 0.0   # Keep output deterministic for classification
      )
    )) %>%
    req_retry(max_tries = 3)  # Automatically retries on HTTP 503

  # Send request; catch network or HTTP errors
  api_response <- tryCatch({
    resp <- req_perform(req)
    resp_body_json(resp)
  }, error = function(e) {
    message(sprintf("API error in batch %d: %s", i, e$message))
    return(NULL)
  })

  # --- Handle the response ---
  batch_success <- FALSE

  if (!is.null(api_response)) {
    raw_text <- api_response$candidates[[1]]$content$parts[[1]]$text

    parsed_response <- tryCatch({
      clean_text <- gsub("```json|```", "", raw_text)
      clean_text <- trimws(clean_text)
      jsonlite::fromJSON(clean_text)
    }, error = function(e) {
      message(sprintf("JSON parse error in batch %d — raw response was malformed.", i))
      return(NULL)
    })

    if (!is.null(parsed_response)) {
      results_list[[as.character(i)]] <- as_tibble(parsed_response)
      batch_success <- TRUE
    }
  }

  # FIX (Issue 3): If the batch failed at any point, log its ID immediately
  # to a file so you have a record without having to scan for NAs afterwards.
  if (!batch_success) {
    failed_batches <- c(failed_batches, i)
    write(
      sprintf("FAILED batch %d at %s", i, Sys.time()),
      file = "failed_batches.log",
      append = TRUE
    )
  }

  # Progress message every 10 batches
  if (i %% 10 == 0) {
    message(sprintf(
      "Processed batch %d of %d (%.1f%%) — %d failed so far",
      i, total_batches, (i / total_batches) * 100, length(failed_batches)
    ))
  }

  # FIX (Issue 4): Save a checkpoint every 15 batches instead of every 50.
  # With ~1,480 batches, saving every 50 means up to 49 batches of lost work
  # after a crash. Every 15 limits that loss to at most 14 batches.
  if (i %% 15 == 0) {
    saveRDS(results_list, "partial_results.rds")
    message(sprintf("Checkpoint saved at batch %d.", i))
  }

  # FIX (Issue 5): Only sleep after a *successful* call.
  # Sleeping after a failure wastes time and gains nothing.
  if (batch_success) Sys.sleep(0.5)
}

# Final checkpoint save
saveRDS(results_list, "partial_results.rds")

# Summary of failures
if (length(failed_batches) > 0) {
  message(sprintf(
    "Done. %d batch(es) failed: %s. See failed_batches.log for details.",
    length(failed_batches), paste(failed_batches, collapse = ", ")
  ))
} else {
  message("Done. All batches completed successfully.")
}
```

#### 5.4 Combine and save results

```{r save-results}
# Combine all successful batch results into one data frame
final_sentiments <- bind_rows(results_list) |> 
  distinct(id, .keep_all = TRUE) #Avoid having duplicates in the results

# Join back to the original data (tweets with no result will show NA)
final_data <- my_data |>  left_join(final_sentiments, by = "id")

write.csv(final_data, "Gemini_results.csv", row.names = FALSE)

message(sprintf(
  "Saved %d rows. Rows with missing sentiment (NA): %d",
  nrow(final_data),
  sum(is.na(final_data$sentiment))
))
```

The output has produced the same number of rows but there are 1355 rows with missing data (NA). This is because some tweets included special characters (like quotation marks), which caused Gemini to generate invalid JSON that couldn't be read or processed.

#### 5.5 Rerun the code on missing batches

```{r retry-failed-batches}
# ── RETRY CHUNK ──────────────────────────────────────────────────────────────
# Runs only the 66 batches that failed in the main loop.
# Safe to re-run multiple times — skips any batch already in results_list.
# ─────────────────────────────────────────────────────────────────────────────

# Load the checkpoint so we don't lose the 1,414 batches that succeeded
results_list <- readRDS("partial_results.rds")

# The 66 failed batch IDs from the log
failed_batches <- c(22, 27, 34, 40, 90, 103, 107, 118, 119, 143, 147, 183,
                    190, 210, 211, 215, 241, 270, 301, 306, 310, 378, 409,
                    416, 488, 537, 679, 709, 724, 744, 747, 760, 766, 809,
                    835, 866, 881, 887, 893, 915, 924, 925, 937, 963, 975,
                    986, 993, 1000, 1022, 1064, 1071, 1082, 1147, 1179, 1192,
                    1202, 1251, 1296, 1301, 1317, 1355, 1365, 1371, 1376,
                    1436, 1454)

still_failed <- c()

for (i in failed_batches) {

  # Skip if already recovered in a previous retry run
  if (as.character(i) %in% names(results_list)) {
    message(sprintf("Batch %d already recovered — skipping.", i))
    next
  }

  current_batch <- my_data %>% filter(batch_id == i)

  # Extra aggressive quote cleaning for problem batches:
  # Replace ALL types of quotes and apostrophes with a simple apostrophe
  current_batch <- current_batch %>%
    mutate(text_clean = str_replace_all(text_clean, '[\u201C\u201D\u201E\u201F\u0022]', "'"),
           text_clean = str_replace_all(text_clean, '[\u2018\u2019\u201A\u201B]', "'"),
           text_clean = str_replace_all(text_clean, '\\\\', ' '))  # remove backslashes

  batch_json_string <- jsonlite::toJSON(
    current_batch %>% select(id, text_clean),
    auto_unbox = TRUE
  )

  prompt <- paste0(
    "You are a strict data processing tool. Do sentiment analysis on the ",
    "text data provided in JSON format below following these instructions:\n\n",
    "1. Sentiment Classification: In English, classify each tweet exactly as ",
    "'positive', 'negative', or 'neutral' (choose only one label per tweet).\n",
    "2. Explanation: Provide a concise and specific explanation for how you ",
    "decided on that label.\n\n",
    "CRITICAL: Return ONLY a valid JSON array of objects containing exactly ",
    "three keys: 'id', 'sentiment', and 'explanation'. ",
    "Do not include markdown formatting, markdown code blocks (like ```json), ",
    "or any conversational filler.\n\n",
    "Data:\n", batch_json_string
  )

  req <- request(url) %>%
    req_body_json(list(
      contents = list(
        list(parts = list(list(text = prompt)))
      ),
      generationConfig = list(temperature = 0.0)
    )) %>%
    req_retry(max_tries = 3)

  api_response <- tryCatch({
    resp <- req_perform(req)
    resp_body_json(resp)
  }, error = function(e) {
    message(sprintf("API error in batch %d: %s", i, e$message))
    return(NULL)
  })

  batch_success <- FALSE

  if (!is.null(api_response)) {
    raw_text <- api_response$candidates[[1]]$content$parts[[1]]$text

    parsed_response <- tryCatch({
      clean_text <- gsub("```json|```", "", raw_text)
      clean_text <- trimws(clean_text)
      jsonlite::fromJSON(clean_text)
    }, error = function(e) {
      message(sprintf("Still failing — batch %d malformed again.", i))
      return(NULL)
    })

    if (!is.null(parsed_response)) {
      results_list[[as.character(i)]] <- as_tibble(parsed_response)
      batch_success <- TRUE
      message(sprintf("✓ Recovered batch %d", i))
    }
  }

  if (!batch_success) {
    still_failed <- c(still_failed, i)
  }

  if (batch_success) Sys.sleep(0.5)
}

# Save updated results
saveRDS(results_list, "partial_results.rds")

# Summary
message(sprintf("\nRetry complete. Recovered %d of %d failed batches.",
                length(failed_batches) - length(still_failed), length(failed_batches)))
if (length(still_failed) > 0) {
  message(sprintf("Still failing after retry: %s", paste(still_failed, collapse = ", ")))
} else {
  message("All previously failed batches recovered successfully!")
}
```

After retrying the code I have recovered 4 of 66 failed batches only. Those batches have tweets so heavily corrupted with special characters that even the aggressive cleaning is not enough.

### 6.Standarisation

#### 6.1 NRC

```{r}
nrc_results <- read.csv("NRC_results.csv", colClasses = c(id = "character"))
```

##### Function to select one single label

```{r}
choose_label <- function(data) {
  if(data$negative > data$positive) {
    label <- "negative"
  } else if (data$positive > data$negative) {
    label <- "positive"
  } else {
    label <- "neutral" # create an extra label, if negative = positive then it is neutral
  }
  return(label)
}
```

##### Run the function on the database

```{r}
nrc_results$nrc_lexicon <- sapply(1:nrow(nrc_results), function(i) choose_label(nrc_results[i, ]))
```

##### Save the new dataframe 

Substitute the old csv with new one, where only a new column is added ("nrc_lexicon")

```{r}
write.csv(nrc_results, "NRC_results.csv", row.names = FALSE)
```

#### 6.2 RoBERTuito

```{r}
robertuito_results <- read.csv("RoBERTuito_results.csv", colClasses = c(id = "character"))
```

##### Change labels

The RoBERTuito model outputs three labels: POS, NEG and NEU. In order to make a comparison between models the labels should be standarised.

```{r}
robertuito_results <- robertuito_results |> 
  mutate(robertuito_transformer = case_when(
    label == "POS" ~ "positive",
    label == "NEG" ~ "negative",
    label == "NEU" ~ "neutral"
  ))
```

##### save the results

```{r}
write.csv(robertuito_results, "RoBERTuito_results.csv", row.names = FALSE)
```


### 7. Join results into one table

### 8. Krippendorff's Alpha 

first we need to pivot the table, each row is a model and column a tweet

install.packages("irr")
library(irr)

then you would do something like this:
```{r}
#alpha_nominal <- kripp.alpha(model_matrix, method = "nominal")
#print(alpha_nominal)
```


