---
title: "Sentiment analysis twitter"
author: "Miriam Lim Leal"
date: "2026-03-26"
output: html_document
---

## Sentiment analysis Twitter

```{r}
library(tidyverse)
library(tidytext)
library(textdata)
library(emoji)
library(readxl)
library(stringi)
library(syuzhet)
library(udpipe)
```

### 1. Load data

```{r}
twitter_data <- read.csv("tweets_political_parties_es.csv", colClasses = c(id = "character"))
```

### 2. Clean tweet text

#### General pipeline 

Remove URLs, mentions and normalise whitespace

```{r}
clean_general <- function(text) {
  
  # Remove URLs
  text <- str_remove_all(text, "https?://\\S+")
  
  # Remove mentions
  text <- str_remove_all(text, "@\\w+")
  
  # Normalize whitespace
  text <- str_squish(text)
  
  return(text)
}
```

```{r}
twitter_transformer <- twitter_data |> 
  mutate(text_clean = map_chr(tweet, clean_general))
```

### 3. Lexicon-based models

#### CSV emoji mapping in Spanish

Load the Spanish emoji mapping (Emoji-SP).

```{r}
emoji_map <- read_excel("emoji_mapping_es.xlsx") |> 
  mutate(definition_lower = str_to_lower(definition)) |> 
  select(emoji, definition_lower, val)
```

#### Function replace emojis with text using Emoji-SP

```{r}
# This replaces the emoji symbol with the Spanish definition from Emoji-SP file
replace_emoji_with_def <- function(text, lexicon) {
  # We use stri_replace_all_fixed to swap emoji symbols for definitions
  # We add spaces and a prefix 'EMOJI_' to help the model distinguish it
  new_text <- stri_replace_all_fixed(
    text, 
    lexicon$emoji, 
    paste0(" ", lexicon$definition_lower, " "), 
    vectorize_all = FALSE
  )
  return(new_text)
}
```

Create a new column with emoji replaced text called emoji_text

```{r}
twitter_lexicon <- twitter_transformer %>%
  mutate(emoji_text = sapply(text_clean, replace_emoji_with_def, lexicon = emoji_map))
```


#### Lexicon-specific cleaning pipeline 

clean data (specifically column emoji_text): drop hashtags symbols, remove punctuation and make everything lowercase.

```{r}
clean_lexicon <- function(text) {
  
  # Keep hashtag words, drop the symbol
  text <- str_replace_all(text, "#(\\w+)", "\\1")
  
  # Remove all punctuation
  text <- str_remove_all(text, "[!\"#$%&'()*+,\\-./:;<=>?@\\[\\\\\\]^_`{|}~¡¿]")
  
  # Lowercase for lexicon matching
  text <- str_to_lower(text)
  
  # Normalize whitespace again after removals
  text <- str_squish(text)
  
  return(text)
}
```


```{r}
twitter_lexicon <- twitter_data |> 
  mutate(
    # Step 1: shared cleaning (URLs, mentions, whitespace)
    text_clean = map_chr(tweet, clean_general),
    
    # Step 2: replace emojis with Spanish definitions (applied to text_clean)
    emoji_text        = map_chr(text_clean, replace_emoji_with_def, lexicon = emoji_map),
    
    # Step 3: lexicon cleaning (hashtags, punctuation, lowercase)
    clean_lexicon_text = map_chr(emoji_text, clean_lexicon)
  ) |>
  select(-emoji_text)    
```

Save this as a csv
```{r}
# write.csv(twitter_lexicon, "twitter_lexicon_and_transformer_clean.csv", row.names = FALSE)
```


#### 1. Lexicon model: NRC

Lemmatize the text function

```{r}
# Function to lemmatize Spanish text
lemmatize_spanish <- function(text, model) {
  # If text is empty, return empty
  if (nchar(text) == 0) return("")
  
  # Annotate the text
  annotation <- udpipe_annotate(model, x = text)
  df <- as.data.frame(annotation)
  
  # Paste the 'lemma' column back together into a single string
  lemmatized_text <- paste(df$lemma, collapse = " ")
  return(lemmatized_text)
}
```

Download the Spanish model

```{r}
# udpipe_download_model(language = "spanish")  # only download once
ud_model <- udpipe_load_model("spanish-gsd-ud-2.5-191206.udpipe")
```

Apply lemmatisation to the text (clean_lexicon_text) REVISE

```{r}
# twitter_lexicon <- twitter_lexicon |> 
#  mutate(lemmatized_text = map_chr(clean_lexicon_text, lemmatize_spanish, model = ud_model))
```

After it has been run once, save to use for other models

```{r}
#write.csv(twitter_lexicon, "twitter_lexicon_lemmatized.csv", row.names = FALSE)
```


Run NRC 

```{r}
nrc_results <- get_nrc_sentiment(twitter_lexicon$clean_lexicon_text, language = "spanish")
```

```{r}
# Bind the id and tweet columns from twitter_lexicon to the NRC results
nrc_results <- bind_cols(
  twitter_lexicon |> select(id, tweet),
  nrc_results
)
```

Save the results of NRC 

```{r}
write.csv(nrc_results, "NRC_results.csv", row.names = FALSE)
```


#### 2. Lexicon model: SEL
