### 5. LLM model: Gemini — ⚠️ TEST VERSION (2 batches only)

This is a **test run** to check that the API call works, measure how long
one batch takes, and estimate the cost before running on the full dataset.

It runs **only 2 batches (= 40 tweets)** and prints a cost estimate at the end.

---

#### 5.1 Setup

```{r libraries}
library(httr2)
library(tidyverse)
library(jsonlite)
```

---

#### 5.2 Prepare data

```{r prepare-data-test}
my_data <- read.csv(
  "twitter_lexicon_and_transformer_clean.csv",
  colClasses = c(id = "character")
)

my_data <- my_data |> select(id, political_party, text_clean)

# Define batch size and assign a batch_id to every row
batch_size <- 20
my_data <- my_data %>%
  mutate(batch_id = ceiling(row_number() / batch_size))

total_batches_full <- max(my_data$batch_id)
message(sprintf("Full dataset has %d batches (%d tweets).", total_batches_full, nrow(my_data)))
```

---

#### 5.3 Test loop — 2 batches only

```{r gemini-test-loop}
# --- API KEY ---
# Make sure GEMINI_API_KEY is in your .Renviron file.
# Run usethis::edit_r_environ() to open it and add:
#   GEMINI_API_KEY=your_actual_key_here
# Then restart R.
API_KEY <- Sys.getenv("GEMINI_KEY")
if (API_KEY == "") stop("GEMINI_KEY not found. Check your .Renviron file.")

# --- MODEL ---
# Using gemini-2.0-flash-lite: cheap, fast, designed for simple classification. 
# IMPORTANT: check model if it is outdated
url <- paste0(
  "https://generativelanguage.googleapis.com/v1beta/models/",
  "gemini-2.5-flash-lite:generateContent?key=", API_KEY
)

# ============================================================
# TEST SETTING: only run batches 1 and 2 (= 40 tweets total).
# Change this to 1:total_batches_full when running for real.
# ============================================================
TEST_BATCHES <- 1:2   # <--- CHANGE THIS LINE TO 1:total_batches_full FOR THE FULL RUN

results_list   <- list()
failed_batches <- c()
token_log      <- list()   # stores token usage per batch for cost estimation

# Record start time to measure total elapsed time
test_start_time <- Sys.time()

for (i in TEST_BATCHES) {

  # Skip if already processed (safety guard against re-running)
  if (as.character(i) %in% names(results_list)) {
    message(sprintf("Batch %d already processed — skipping.", i))
    next
  }

  current_batch <- my_data %>% filter(batch_id == i)

  # toJSON handles all escaping — do NOT manually escape quotes before this
  batch_json_string <- jsonlite::toJSON(
    current_batch %>% select(id, text_clean),
    auto_unbox = TRUE
  )

  prompt <- paste0(
    "You are a strict data processing tool. Do sentiment analysis on the ",
    "text data provided in JSON format below following these instructions:\n\n",
    "1. Sentiment Classification: In English, classify each tweet exactly as ",
    "'positive', 'negative', or 'neutral' (choose only one label per tweet).\n",
    "2. Explanation: Provide a concise and specific explanation for how you ",
    "decided on that label.\n\n",
    "CRITICAL: Return ONLY a valid JSON array of objects containing exactly ",
    "three keys: 'id', 'sentiment', and 'explanation'. ",
    "Do not include markdown formatting, markdown code blocks (like ```json), ",
    "or any conversational filler.\n\n",
    "Data:\n", batch_json_string
  )

  req <- request(url) %>%
    req_body_json(list(
      contents = list(
        list(parts = list(list(text = prompt)))
      ),
      generationConfig = list(
        temperature = 0.0
      )
    )) %>%
    req_retry(max_tries = 3)

  # Time each individual batch call
  batch_start <- Sys.time()

  api_response <- tryCatch({
    resp <- req_perform(req)
    resp_body_json(resp)
  }, error = function(e) {
    message(sprintf("API error in batch %d: %s", i, e$message))
    return(NULL)
  })

  batch_elapsed <- as.numeric(Sys.time() - batch_start, units = "secs")

  batch_success <- FALSE

  if (!is.null(api_response)) {

    # --- Extract token usage from the API response ---
    # Google returns token counts in usageMetadata — log them for cost estimate
    usage <- api_response$usageMetadata
    input_tokens  <- usage$promptTokenCount
    output_tokens <- usage$candidatesTokenCount
    token_log[[as.character(i)]] <- list(
      input  = input_tokens,
      output = output_tokens
    )
    message(sprintf(
      "Batch %d — %.1f sec | input tokens: %d | output tokens: %d",
      i, batch_elapsed, input_tokens, output_tokens
    ))

    raw_text <- api_response$candidates[[1]]$content$parts[[1]]$text

    parsed_response <- tryCatch({
      clean_text <- gsub("```json|```", "", raw_text)
      clean_text <- trimws(clean_text)
      jsonlite::fromJSON(clean_text)
    }, error = function(e) {
      message(sprintf("JSON parse error in batch %d — raw response was malformed.", i))
      return(NULL)
    })

    if (!is.null(parsed_response)) {
      results_list[[as.character(i)]] <- as_tibble(parsed_response)
      batch_success <- TRUE
    }
  } else {
    # Log failure
    failed_batches <- c(failed_batches, i)
    message(sprintf("Batch %d FAILED.", i))
  }

  # Only sleep between successful calls, not after failures
  if (batch_success) Sys.sleep(0.5)
}

# -------------------------------------------------------
# SUMMARY: time + cost estimate for the full dataset
# -------------------------------------------------------
test_elapsed <- as.numeric(Sys.time() - test_start_time, units = "secs")
n_test       <- length(TEST_BATCHES)
total_batches_full <- max(my_data$batch_id)

# Average tokens per batch from the test
avg_input  <- mean(sapply(token_log, `[[`, "input"))
avg_output <- mean(sapply(token_log, `[[`, "output"))

# gemini-2.0-flash-lite pricing (as of 2025):
#   Input:  $0.015 per 1 million tokens
#   Output: $0.060 per 1 million tokens
price_input_per_million  <- 0.015
price_output_per_million <- 0.060

total_input_tokens  <- avg_input  * total_batches_full
total_output_tokens <- avg_output * total_batches_full

cost_input  <- (total_input_tokens  / 1e6) * price_input_per_million
cost_output <- (total_output_tokens / 1e6) * price_output_per_million
cost_total_usd <- cost_input + cost_output
cost_total_eur <- cost_total_usd * 0.92  # approximate USD -> EUR

# Estimated time for full run (seconds per batch * total batches)
avg_secs_per_batch    <- test_elapsed / n_test
est_total_secs        <- avg_secs_per_batch * total_batches_full
est_total_mins        <- est_total_secs / 60

message("\n========== TEST RESULTS ==========")
message(sprintf("Test batches run:         %d", n_test))
message(sprintf("Tweets processed:         %d", n_test * batch_size))
message(sprintf("Time for test:            %.1f seconds", test_elapsed))
message(sprintf("Avg time per batch:       %.1f seconds", avg_secs_per_batch))
message(sprintf("Avg input tokens/batch:   %.0f", avg_input))
message(sprintf("Avg output tokens/batch:  %.0f", avg_output))
message("-----------------------------------")
message(sprintf("FULL dataset batches:     %d", total_batches_full))
message(sprintf("Estimated total time:     %.0f minutes", est_total_mins))
message(sprintf("Estimated total cost:     $%.4f USD (~€%.4f EUR)", cost_total_usd, cost_total_eur))
message("===================================")

if (length(failed_batches) > 0) {
  message(sprintf("WARNING: %d batch(es) failed: %s", length(failed_batches), paste(failed_batches, collapse = ", ")))
}
```

---

#### 5.4 Inspect test results

```{r inspect-test-results}
# Check that the results look correct before running the full dataset
test_results <- bind_rows(results_list)
print(test_results)
```
